The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-12-15 manuscript.Rmd
refactor with using postgis queries everywhere

Marine invertebrates are underrepresented in iDigBio and OBIS. This underrepresentation is manifest in the small fraction of records that marine invertebrates represent. Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$usa_no_fossil*100)`% of the US records (`r round(idigbio_stats$usa_no_fossil/1e6, 1)` million). Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).

@@

n_spp_gom <- combined_records %>%
    dplyr::filter(is_in_gom == TRUE) %>%
    count_spp

@@

prop_idigbio_continental_shelf <-
   mean(idigbio_us_inverts_records_all_with_depth$marmap_depth < 0 &
        idigbio_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
   format_output()

prop_obis_continental_shelf <-
    mean(obis_us_inverts_records_all_with_depth$marmap_depth < 0 &
         obis_us_inverts_records_all_with_depth$marmap_depth > -200) * 100 %>%
    format_output()
```

Most samples have been collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental shelf`% (OBIS) of the records collected between 0-200m. More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_in_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively). The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).

@@

Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters. In our dataset, we identified at least `r format(sum(idigbio_mms_summary$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985. Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name))` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.

@@

We extracted `r format(sum(gom_worms$is_marine, na.rm=TRUE), big.mark=",")` and `r format(sum(kozloff_worms$is_marine, na.rm=TRUE), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively. For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.

@@

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within US waters. Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems. Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific have been digitized.

@@

This study is fully reproducible and the code needed to perform the analyses included here is available from: https://github.com/fmichonneau/sok-marine-biodiversity, and archived on Zenodo (doi:xxx.xxx.xxxx). The data downloaded from the databases at the time of analysis are archived on Zenodo (doi:xxx.xxx.xxx)/

>>>>

Marine invertebrates are underrepresented in iDigBio and OBIS. This underrepresentation is manifest in the small fraction of records that marine invertebrates represent. Extant marine invertebrates occurring in the US EEZ represent  `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million). Marine invertebrates however represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]).

@@

n_spp_gom <- combined_records %>%
    dplyr::filter(within_gom == TRUE) %>%
    count_spp

@@

prop_idigbio_continental_shelf <- (
   mean(idigbio_us_inverts_records_all_with_depth$marmap_depth < 0 &
        idigbio_us_inverts_records_all_with_depth$marmap_depth > -200) * 100) %>%
   format_output()

prop_obis_continental_shelf <- (
    mean(obis_us_inverts_records_all_with_depth$marmap_depth < 0 &
         obis_us_inverts_records_all_with_depth$marmap_depth > -200) * 100) %>%
    format_output()
```

Most samples have been collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m. More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$is_within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively). The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species).

@@

Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters. In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985. Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name))` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.

@@

We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively. For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.

@@

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_eez*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_eez, big.mark=",")`) fall within US waters. Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems. Similarly, only a small fraction of the marine invertebrates records from instiutions on the Pacific has been digitized.

@@

This study is fully reproducible and the code needed to perform the analyses included here is available from: https://github.com/fmichonneau/sok-marine-biodiversity, and archived on Zenodo (doi:xxx.xxx.xxxx). The data downloaded from the databases at the time of analysis are archived on Zenodo (doi:xxx.xxx.xxx).
